---
title: "Youth & Adult Literacy Rates from 2010 to 2018"
author: "Samina Issa - Fall 2023"
date: "2023-11-29"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
```

# Introduction

[Youth & Adult Literacy Rates Datasets Link](https://www.kaggle.com/datasets/thedevastator/youth-adult-literacy-rates-in-2019/data)


The data is a CSV from Kaggle. This dataset contains data on literacy rates for youth and adults in various regions and countries.  

# Questions:
1. 
2. 
3. 
4.

## Why Literacy Project?

Growing up, many people in my community, neighborhood and family were not able to read and write. So I wanted to see the literacy rate in recent years and picked it for the project

# Library Loading 
- library(dplyr): using for data manipulation

- library(tidyverse): using to transform and better present data

- library(ggplot2): using for data visualization and plotting charts

- library(plotly): using for creating interactive web-based graphs 

```{r message = FALSE}
#install.packages("tidyverse")

#using this for data manipulation
library(dplyr)
#using this to tidy up the data
library(tidyverse)
#using for visual data
library(ggplot2)
#using for interactive graphs
library(plotly)


```

# Reading in Data 

```{r message = FALSE}
# Read CSV file from a subfolder

LIT_rate_data <- read_csv("Literacy rates.csv")
glimpse(LIT_rate_data)

LITdata <- read_csv("Adults_15YrsAndUp.csv")
glimpse(LITdata)


#LIT_rate_data %>% arrange(desc(Year), desc(Country))

```
# Countries with Higest Literacy rate

```{r message = FALSE}

LIT_rate_data %>% 
  group_by(Country) %>% 
  summarise(mean_lit_rate = mean(`Literacy rate`)) %>% arrange(desc(mean_lit_rate))




```
# Countries with Lowest Literacy rate

```{r message = FALSE}

LIT_rate_data %>% 
  group_by(Country) %>% 
  summarise(mean_lit_rate = mean(`Literacy rate`)) %>%
  arrange(mean_lit_rate)


```


# Countries with High Female Literacy rate

```{r message = FALSE}

LITdata %>% mutate(diff = Male - Female) %>% arrange(diff) %>% select(Country, Male, Female, diff)

```
# Countries with Low Female Literacy rate

```{r message = FALSE}

LITdata %>% mutate(diff = Male - Female) %>% arrange(-diff) %>% select(Country, Male, Female, diff)



```

# Literacy rate by Age

```{r message = FALSE}

p <- ggplot(data = LIT_rate_data) + 
  geom_bar(mapping = aes(x = Year, fill = Age), position = "dodge")

ggplotly(p)

p <- ggplot(data = LIT_rate_data) + 
  geom_col(mapping = aes(x = Age, y = `Literacy rate`, fill = Age))

ggplotly(p)

```

# Male & Female Literacy rate by region


```{r message = FALSE}
#Literacy rate Country Table
#table(LIT_rate_data$Country)


# rate for female in each region
f <- LIT_rate_data %>% 
        filter(Gender == "female") %>% 
        ggplot(aes(x = reorder(Region, `Literacy rate`), y = `Literacy rate`))

f + 
  geom_col() + 
  coord_flip() + labs(title = "Female Literacy rate by region")

# rate for male in each region
m <- LIT_rate_data %>% 
        filter(Gender == "male") %>% 
        ggplot(aes(x = reorder(Region, `Literacy rate`), y = `Literacy rate`))+ 
        geom_col() + 
        coord_flip() + labs(title = "Male Literacy rate by region")

m


LIT_rate_data %>% 
  group_by(Region) %>% 
  summarise(mean_lit_rate = mean(`Literacy rate`)) %>%
  arrange(-mean_lit_rate)


```


# Measuring the progress of a Region's literacy rate over time

```{r message = FALSE}

p <- LIT_rate_data %>%
  group_by(Region, Year) %>%
  summarise(mean_literacy = mean(`Literacy rate`, na.rm = TRUE)) %>%
  ggplot(aes(x = Year, y = mean_literacy, color = Region)) +
  geom_line() +
  facet_wrap(~ Region)

ggplotly(p)

```

# Region's literacy rate by Gender
```{r message = FALSE}
p <- ggplot(data = LIT_rate_data %>%
         filter(Gender %in% c("male", "female"))) + 
  geom_bar(mapping = aes(x = Region, fill = Gender))+ coord_flip()

ggplotly(p)


```
# ---

```{r message = FALSE}
ggplot(LIT_rate_data %>% filter(Gender == "female"), aes(x = Year, y = `Literacy rate`)) + geom_line() + geom_point()
```

# ...Measuring the progress of a country's literacy rate over time

```{r message = FALSE}
jgender <- c("female", "male")
ggplot(LIT_rate_data %>% filter(Gender %in% jgender) %>% group_by(Year) %>% summarise(Lit_mean = mean(na.omit((`Literacy rate`)))), aes(x = Year, y = Lit_mean)) +
        geom_line() + geom_point()

p <- ggplot(LIT_rate_data %>% filter(Gender == "female") %>% group_by(Year) %>% summarise(Lit_mean = mean(na.omit(`Literacy rate`))), aes(x = Year, y = Lit_mean)) + geom_line() + geom_point()
ggplotly(p)

```

# Comparing the literacy rates of different countries


```{r message = FALSE}

```

# Comparing the literacy rates of different age groups in a country

```{r message = FALSE}

```

# Analysing the literacy rates of a country


```{r message = FALSE}
library(plotly)
p <- ggplot(LIT_rate_data %>% filter(Gender == "male") %>% group_by(Year, Region) %>% summarise(Lit_mean = mean(na.omit(`Literacy rate`))), aes(x = Year, y = Lit_mean, color = Region)) + geom_line() + geom_point()
ggplotly(p)

p <- ggplot(LIT_rate_data %>% filter(Gender == "female", Region == "Sub-Saharan Africa") %>% group_by(Year, Country) %>% summarise(Lit_mean = mean(na.omit(`Literacy rate`))), aes(x = Year, y = Lit_mean, color = Country)) + geom_line() + geom_point()
ggplotly(p)
```

# Summary


```{r message = FALSE}

```



```{r message = FALSE}

```



```{r message = FALSE}

```



```{r message = FALSE}

```


```{r message = FALSE}

```



```{r message = FALSE}

```
