---
title: "Youth & Adult Literacy Rates from 2010 to 2018"
author: "Samina Issa - Fall 2023"
date: "2023-11-29"
output: slidy_presentation
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
```

# Introduction

[Youth & Adult Literacy Rates Datasets Link](https://www.kaggle.com/datasets/thedevastator/youth-adult-literacy-rates-in-2019/data)


The data is a CSV from Kaggle. This dataset contains data on literacy rates for youth and adults in various regions and countries.  It has two csv files. 

The dataset can be used to help policy-makers understand where interventions are needed to improve literacy rates



# Analysing:
* Comparing the literacy rates of different 
  - regions
  - countries
  - age groups
  - genders


## Why Literacy Project?

Growing up, many people in my community, neighborhood and family were not able to read and write. So I wanted to see the literacy rate in recent years and picked it for the project

# Library Loading 
- **library(dplyr):** using for data manipulation

- **library(tidyverse):** using to transform and better present data

- **library(ggplot2):** using for data visualization and plotting charts

- **library(plotly):** using for creating interactive web-based graphs 

- **library(knitr):** using for formatting tables 

- **library(kableExtra):** using for formatting tables and manipulate table styles 

```{r message = FALSE}
#install.packages("tidyverse")
#install.packages("knitr")

#using this for data manipulation
library(dplyr)
#using this to tidy up the data
library(tidyverse)
#using for visual data
library(ggplot2)
#using for interactive graphs
library(plotly)
library(kableExtra)
library(knitr)
```

# Reading in Data 
```{r message = FALSE}
# Read CSV file from a subfolder

LIT_rate_data <- read_csv("Literacy rates.csv")
glimpse(LIT_rate_data)

LITdata <- read_csv("Adults_15YrsAndUp.csv")
glimpse(LITdata)

```

# Countries with Higest Literacy rate
```{r message = FALSE}

LIT_rate_data %>% 
  group_by(Country) %>% 
  summarise(mean_lit_rate = mean(`Literacy rate`)) %>% 
  arrange(desc(mean_lit_rate)) %>% 
  slice(1:10) %>%
  rename("Country Name" = Country, "Mean Literacy Rate" = mean_lit_rate) %>%
  kable("html", digits = 3) %>%
  kable_styling(full_width = FALSE, bootstrap_options = "striped", position = "left") 

```

# Countries with Lowest Literacy rate
```{r message = FALSE}

LIT_rate_data %>% 
  group_by(Country) %>% 
  summarise(mean_lit_rate = mean(`Literacy rate`)) %>%
  arrange(mean_lit_rate) %>% 
  slice(1:10) %>%
  rename("Country Name" = Country, "Mean Literacy Rate" = mean_lit_rate) %>%
  kable("html", digits = 3) %>%
  kable_styling(full_width = FALSE, bootstrap_options = "striped", position = "left")


```


# Literacy rates of different age groups by year
```{r message = FALSE}

LIT_rate_data %>%
  group_by(Age, Year) %>% 
  summarize(mean_lit_rate = mean(`Literacy rate`)) %>% 
  ggplot() + 
  geom_col(mapping = aes(y  = mean_lit_rate, x = Year, fill = Age), position = "dodge") + 
  labs(title = "Literacy rate by Year", x = "Year", y = "Mean Literacy Rate")

```

# Mean literacy rates by age group
```{r message = FALSE}
 

p <- ggplot(data = LIT_rate_data) +
  geom_col(mapping = aes(x = Age, y = `Literacy rate`, fill = Age)) + 
  labs(title = "Average literacy rate by Age", x = "Age", y = "Literacy rate")

ggplotly(p)

```


# Progress of a Region's literacy rate over time
```{r message = FALSE}
library(plotly)
p <- ggplot(LIT_rate_data %>% 
              filter(Gender == "male") %>% 
              group_by(Year, Region) %>% 
              summarise(Lit_mean = mean(na.omit(`Literacy rate`))), 
            aes(x = Year, y = Lit_mean, color = Region)) + 
  geom_line() + 
  geom_point()+ 
  labs(title = "Region's literacy rate over time", x = "Year", y = "Mean Literacy rate")

ggplotly(p)

```


# Progress of a Region's literacy rate over time
```{r message = FALSE, fig.width=12}

p <- LIT_rate_data %>%
  group_by(Region, Year) %>%
  summarise(mean_literacy = mean(`Literacy rate`, na.rm = TRUE)) %>%
  ggplot(aes(x = Year, y = mean_literacy, color = Region)) +
  geom_line() +
  labs(title = "Region's literacy rate over time", x = "Year", y = "Mean Literacy rate") +
  facet_wrap(~ Region)

ggplotly(p)

```


# Literacy rates by country of Sub-Saharan Africa Region
```{r message = FALSE}
library(plotly)

p <- ggplot(LIT_rate_data %>% 
              filter(Gender == "female", Region == "Sub-Saharan Africa") %>% 
              group_by(Year, Country) %>% 
              summarise(Lit_mean = mean(na.omit(`Literacy rate`))), 
            aes(x = Year, y = Lit_mean, color = Country)) + 
  geom_line() + 
  geom_point() +
  labs(title = "Sub-Saharan Africa Region's literacy rate over time", x = "Year", y = "Mean Literacy rate")

ggplotly(p)
```

# Sub-Saharan Africa Region
```{r message = FALSE}

LIT_rate_data %>%
  filter(Region == "Sub-Saharan Africa") %>%
  group_by(Country, Year) %>%
  summarise(mean_lit_rate = mean(`Literacy rate`)) %>%
  rename("Country Name" = Country, "Mean Literacy Rate" = mean_lit_rate) %>%
  kable("html", digits = 3) %>%
  kable_styling(full_width = FALSE, bootstrap_options = "striped", position = "left") %>%
  add_header_above(c(" " = 1, "Literacy Rate Analysis" = 2))

```

# Literacy rates of Year 2011 & 2013 of Sub-Saharan Africa Region
```{r message = FALSE}


LIT_rate_data %>% 
  filter(Region == "Sub-Saharan Africa", Year == 2011) %>% 
  group_by(Country, Year) %>% 
  summarise(mean_lit_rate = mean(`Literacy rate`)) %>%
  rename("Country Name" = Country, "Mean Literacy Rate" = mean_lit_rate) %>%
  kable("html", digits = 3) %>%
  kable_styling(full_width = FALSE, bootstrap_options = "striped", position = "left")



LIT_rate_data %>% 
  filter(Region == "Sub-Saharan Africa", Year == 2013) %>% 
  group_by(Country, Year) %>% 
  summarise(mean_lit_rate = mean(`Literacy rate`)) %>%
  rename("Country Name" = Country, "Mean Literacy Rate" = mean_lit_rate) %>%
  kable("html", digits = 3) %>%
  kable_styling(full_width = FALSE, bootstrap_options = "striped", position = "left")



```

# Literacy rates of countries of Eastern and South-Eastern Asia Region
```{r message = FALSE}


p <- ggplot(LIT_rate_data %>% 
              filter(Gender == "female", Region == "Eastern and South-Eastern Asia") %>% 
              group_by(Year, Country) %>% 
              summarise(Lit_mean = mean(na.omit(`Literacy rate`))), 
            aes(x = Year, y = Lit_mean, color = Country)) + 
  geom_line() + 
  geom_point() +
  labs(title = "Eastern and South-Eastern Asia Region's literacy rate over time", x = "Year", y = "Mean Literacy rate")

ggplotly(p)
```

# Eastern and South-Eastern Asia Region
```{r message = FALSE}


LIT_rate_data %>% 
  filter(Region == "Eastern and South-Eastern Asia") %>% 
  group_by(Country, Year) %>% 
  summarise(mean_lit_rate = mean(`Literacy rate`)) %>%
  rename("Country Name" = Country, "Mean Literacy Rate" = mean_lit_rate) %>%
  kable("html", digits = 3) %>%
  kable_styling(full_width = FALSE, bootstrap_options = "striped", position = "left") %>%
  add_header_above(c(" " = 1, "Literacy Rate Analysis" = 2))

```


# Male & Female Literacy rate by region
```{r message = FALSE}

LIT_rate_data %>%
  group_by(Region, Gender) %>%
  summarise(mean_lit_rate = mean(`Literacy rate`), .groups = "drop") %>%
  pivot_wider(names_from = Gender, values_from = mean_lit_rate) %>%
  mutate(across(c(male, female), ~ round(., 3))) %>%
  select(Region, male, female) %>%
  arrange(Region) %>%
  rename("Male" = male, "Female" = female) %>%
  kable("html", digits = 3) %>%
  kable_styling(full_width = FALSE, bootstrap_options = "striped", position = "left")

LIT_rate_data %>% 
  filter(Gender %in% c("male", "female")) %>% 
  group_by(Gender, Region) %>% 
  summarise(mean_lit_rate = mean(`Literacy rate`)) %>% 
  ggplot() +
  geom_col(mapping = aes(x = reorder(Region, mean_lit_rate), y = mean_lit_rate, fill = Gender), position = "dodge")+ 
  coord_flip() +
  labs(title = "Mean literacy rate by Gender", x = "Region", y = "Mean Literacy rate")


```

# Countries with High Female Literacy rate
```{r message = FALSE}

LITdata %>% 
  mutate(diff = Male - Female) %>% 
  arrange(diff) %>% 
  select(Country, Male, Female, diff) %>% 
  slice(1:10) %>%
  rename("Difference" = diff) %>%
  kable() %>%
  kable_styling(full_width = FALSE, bootstrap_options = "striped", position = "center")

```


# Countries with Low Female Literacy rate
```{r message = FALSE}
LITdata %>% 
  mutate(diff = Male - Female) %>% 
  arrange(-diff) %>% 
  select(Country, Male, Female, diff) %>% 
  slice(1:10) %>%
  rename("Difference" = diff) %>%
  kable() %>%
  kable_styling(full_width = FALSE, bootstrap_options = "striped", position = "center")

```


# Male & Female Literacy rate over time

```{r message = FALSE}

p <- LIT_rate_data %>% 
  filter(Gender %in% c( "female", "male")) %>% 
  group_by(Gender, Year) %>% 
  summarise(Lit_mean = mean(na.omit(`Literacy rate`))) %>%  
  ggplot(aes(x = Year, y = Lit_mean, color = Gender)) + 
  geom_line() + 
  geom_point() +
  labs(title = "Mean literacy rate by Gender", x = "Year", y = "Mean Literacy rate")

ggplotly(p)

```




# Summary
- Gap between male and female is consistent over the years
- More programs are required for females  &
- 65+ age group 
- Data collection is inconsistent from year to year 
- Need more data to compare the regions or countries
